 Casting multiple shadows: interactive data visualisation with tours and embeddings

### 2020-02-07 paper.Rmd
adding citations
re wrap
add section on design

Non-linear methods
generally perform preprocessing on the high-dimensional data such as generating
a neighborhood graph and perform transformations on the preprocessed form. We
restrict our attention to three methods that are commonly used in
high-throughput biology: t-distributed stochastic neighbor embedding (t-SNE), 
uniform manifold alignment and projection (UMAP), and potential of heat-diffusion for affinity-based transition embedding (PHATE). The t-SNE algorithm estimates the similarity of (Euclidean) distances of points in a high dimensional space using a Gaussian distribution and then  
estimates a configuration in the low dimensional embedding space by modelling
similarities using a t-distribution with 1 degree of freedom. The resulting
configuration is the one that minimizes the Kulback-Leibler divergence
between the two distributions. A recent theoretical contribution by @Linderman2019-dq proved that t-SNE can recover spherical and well separated 
cluster shapes, and proposed new approaches for tuning the optimisation parameters. It is a known problem that t-SNE can have trouble recovering
global structure and that configurations can be highly dependent on
how the algorithm is initialised and parameterised 
[@Wattenberg2016-ji; @Kobak2019-lm]. UMAP is a method that is
related to LargeVis, and like t-SNE acts on the k-nearest neighbor graph. Its
main differences are that it uses a different cost function (cross entropy) which is optimized using stochastic gradient descent and defines a different
kernel for similiarites in the low dimensional space. Due to it's computational
speed it's possible to generate UMAP embeddings in more than three dimensions. Finally, PHATE, inspired by diffusion maps, is based on estimating an affinity matrix via a distance matrix and k-nearest neighbors graph. The algorithm
denoises estimated distances in high dimensional space via transforming the
affinity matrix into a Markov transition probaility matrix and diffusing this 
matrix over a fixed number of time steps. Then the diffused probabilities
are transformed once more to construct a distance matrix, and multidimensional
scaling is performed to construct a 2d embedding for visualization.

@@

## Multiple views, multiple contexts

## Interactions and visual analytics for embeddings

>>>>

Non-linear methods generally perform preprocessing on the high-dimensional data
such as generating a neighborhood graph and perform transformations on the
preprocessed form. We restrict our attention to three methods that are commonly
used in high-throughput biology: t-distributed stochastic neighbor embedding
(t-SNE), uniform manifold alignment and projection (UMAP), and potential of
heat-diffusion for affinity-based transition embedding (PHATE). The t-SNE
algorithm estimates the similarity of (Euclidean) distances of points in a high
dimensional space using a Gaussian distribution and then estimates a
configuration in the low dimensional embedding space by modelling similarities
using a t-distribution with 1 degree of freedom. The resulting configuration is
the one that minimizes the Kulback-Leibler divergence between the two
distributions. A recent theoretical contribution by @Linderman2019-dq proved
that t-SNE can recover spherical and well separated cluster shapes, and
proposed new approaches for tuning the optimisation parameters. It is a known
problem that t-SNE can have trouble recovering global structure and that
configurations can be highly dependent on how the algorithm is initialised and
parameterised [@Wattenberg2016-ji; @Kobak2019-lm]. UMAP is a method that is
related to LargeVis [@Tang2016-oz], and like t-SNE acts on the k-nearest
neighbor graph. Its main differences are that it uses a different cost function
(cross entropy) which is optimized using stochastic gradient descent and
defines a different kernel for similiarites in the low dimensional space. Due
to it's computational speed it's possible to generate UMAP embeddings in more
than three dimensions. Finally, PHATE, inspired by diffusion maps, is based on
estimating an affinity matrix via a distance matrix and k-nearest neighbors
graph. The algorithm denoises estimated distances in high dimensional space via
transforming the affinity matrix into a Markov transition probaility matrix and
diffusing this matrix over a fixed number of time steps. Then the diffused
probabilities are transformed once more to construct a distance matrix, and
multidimensional scaling is performed to construct a 2d embedding for
visualization.

@@

We propose using tours as part of an analyst's workflow in in performing
dimensionality reduction tasks. We have made extensive use of ensemble
graphics, that is aligning related plots alongside each other to provide
context. As we will see in the case studies, this allows analysts to quickly
compare views from embedding methods and allows them to see how the embedding
method alters the global structure of their data. Using ensembles allows the
use of interaction techniques, that allow analysts to perform cluster
orientation tasks via linking multiple views. This approach allows our
interface, to achieve the three principles for interactive high-dimensional
data visualisation outlined by @Buja1996-fk: finding gestalt, posing queries,
and making comparisons.

## Finding Gestalt: focus and context

To investigate latent structure and the shape of a high dimensional dataset, a
tour can be run without the use of an external embedding. It is often useful to
first run principal components on the input as an intial dimension reduction
step, and then tour a subset of those components instead, i.e. by selecting
them from a scree plot. The default tour layout is a scatter plot with an axis
layout displaying the magnitude and direction of each basis vector. Since the
tour is dynamic, it is often useful to be able to pause and highlight a
particular view. In our interface, brushing will pause the tour path, allowing
users to identify 'interesting' projections. The domain of the axis scales from
running a tour is called the half range, and is computed by rescaling the input
data onto hyperdimensional unit cube. We bind the half range to a mouse wheel
event, allowing a user to pan and zoom on the tour view dynamically. This is
useful for peeling back dense clumps of points to reveal structure.

## Posing Queries: multiple views, multiple contexts

## Making comparsions
